---
title: "Final Project 1"
author: "Parker Lambert"
date: "2023-07-24"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Name: Parker Lambert

Class: 103

Assignment: Final Project 1

Due: 7/25/23

# Final Project 1

### Libraries

```{r, results='hide'}
library(tidyverse)
library(ggplot2)
library(tibble)
library(hrbrthemes)
```

### Import Data and Wrangle

```{r}
# importing data from file
GE <- read.csv("QBS103_finalProject_geneExpression.csv")
#summary(GE)
MD <- read.csv("QBS103_finalProject_metadata.csv")
# transpose the GE data table
TGE <- setNames(data.frame(t(GE[,-1])),GE[,1])
TGE2 <- tibble::rownames_to_column(TGE,'participant_id')
# linking the two data tables
df <- full_join(TGE2,MD,by = "participant_id")
#df
```

## Identify from Data

### One Gene

The gene I have selected is `ABCF2` official name is: `ATP binding cassette subfamily F member 2`

It is a member of the ATP-Binding cassette transporter super-family. It helps with transport of various molecules across the extra- and intracellular membranes. Alterations in this gene may be involved in cancer progression.

*All information came from the: [National Library of Medicine](https://www.ncbi.nlm.nih.gov/gene/10061)*

### Continuous Covariate

The continuous covariate I selected is `ferritin.ng.ml`

### Two Categorical Covariates

The two categorical covariates are `sex` and `mechanical_ventilation`

## Histogram

```{r, warning=FALSE}
# select gene
ggplot(df,aes(df$ABCF2)) +
  # choose histogram
  geom_histogram(binwidth = 1) +
  # add labels
  labs(title = 'Gene Expression Histogram',y= 'Frequency',x='ABCF2 Expression') +
  # make pretty
  theme_ipsum()
```

## Scatterplot

```{r,warning=FALSE}
# select desired collumn
df2 <- df |>
  select(ferritin.ng.ml.,ABCF2)
# make ferritin an integer
df2$ferritin.ng.ml. <- as.integer(df2$ferritin.ng.ml.)
# remove na
#df2 |> 
#  na.omit()
#summary(df2)
```

```{r, warning=FALSE}
# select collumns 
ggplot(df2,aes(df2$ABCF2,df2$ferritin.ng.ml.)) +
  # choose scatterplot
  geom_point() +
  #label the plot
  labs(title = 'Gene ABCF2 by Ferritin Level',subtitle = 'Ferritin Normal Range is 24 to 336 ng/ml', x = 'Gene ANCF2', y = 'Ferritin ng/ml') +
  # make pretty by adding theme
  theme_ipsum()
```

## Box Plot

```{r}
# make sex a factor
df$sex<-factor(df$sex)
# remove the unknwon
dff <- df[df$sex != " unknown",]
# refactor sex and save to new df
dff$sex <- factor(dff$sex, levels = c(' male', ' female'))
# make mechanical_ventilation a factor
dff$mechanical_ventilation <- factor(dff$mechanical_ventilation)
#dff$mechanical_ventilation
```

```{r}
# select columns
ggplot(data = na.omit(dff[,c('ABCF2','sex','mechanical_ventilation')]),aes(sex,ABCF2,color=mechanical_ventilation)) +
  # choose boxplot
  geom_boxplot() +
  # create labels
  labs(title = 'Gene expression by Sex and Ventilation',x = 'Sex', y = 'Gene ABCF2', color = 'Mechanical Ventilation') +
  # select theme
  theme_ipsum()
```
